
Explore and Summarize Data by Germanas Latvaitis
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)

library(gridExtra)

library(knitr)

library(GGally)

library(dplyr)

```

```{r echo=FALSE, Load_the_Data}

wine_data <- read.csv('wineQualityReds.csv',header = T)

```

#Initial data exploration

```{r echo=FALSE}

summary(wine_data)
str(wine_data)

```

# Univariate Plots Section

First I will check the distribution of my featured variable: quality.

```{r echo=FALSE}
ggplot(aes(x = quality), data = wine_data, binwidth = 1) +
  geom_bar(fill="white",colour="black")

```

```{r echo=FALSE,message=FALSE, warning=FALSE}
summary(wine_data$quality)
```

In this histogram I can see that most of the values are binned in quality 
range of 5 - 6. To see the distribution better I will transform the plot 
using scale_y_log10.

```{r echo=FALSE,message=FALSE, warning=FALSE}
ggplot(aes(x = quality), data = wine_data) + 
  geom_bar() +
  scale_y_log10()
```

Transforming the plot with log10 shows me a normal distribution.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol), data = wine_data) +
  geom_histogram(fill="white",colour="black")

```

```{r echo=FALSE}

summary(wine_data$alcohol)

```

With this histogram I wanted to see which alcohol consistensy is the most 
popular. I can now see that there are some outliers which I will try to adress
later.

Now I will draw a histogram of violatile.acidity distribution:

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity), data = wine_data) +
  geom_histogram(fill="white",colour="black")

```
```{r echo=FALSE}

summary(wine_data$volatile.acidity)

```
According to summary, the 3rd quartile is 0.64 and the max value is 1.58 which
tells me that there is a possibility of outlier here.

Finaly I will check the distribution of pH levels

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = pH), data = wine_data) +
  geom_histogram(fill="white",colour="black")

```

```{r echo=FALSE}

summary(wine_data$pH)

```

This looks like a normal distribution to me. With min value 2.7 and max value 4.

```{r echo=FALSE}

summary(subset(wine_data, quality == 8))

```

Now I checked the most popular alcohol proof and with the highest quality which
is 8. After ploting I gues that the quality doesn't really depend purely on the 
alcohol concentration. The plot looks similar to uniform distribution plot.

```{r echo=FALSE, message=FALSE, warning=FALSE}

qplot(data = subset(wine_data, quality == 8), x = alcohol) +
  geom_histogram(fill="white",colour="black")

```
```{r echo=FALSE, message=FALSE, warning=FALSE}

alcohol_high_quality <- subset(wine_data$alcohol, wine_data$quality == 8)

```
```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(alcohol_high_quality)

```

The highest quality wine has the minimum alcohol percentage of 9.8 and maximum 
of 14.

```{r echo=FALSE, message=FALSE, warning=FALSE}
plot1 <- qplot(data = subset(wine_data, quality > 5), x = alcohol) +
  geom_histogram(fill="white",colour="black") +
  labs(list(title = "Wines with quality lower than 5", x = "Alcohol percentage",
            y = "Count")) +
  coord_cartesian(ylim = c(0, 200))

plot2 <- qplot(data = subset(wine_data, quality <= 5), x = alcohol) +
  geom_histogram(fill="white",colour="black") + 
  labs(list(title= "Wines with quality greater than 5", x= "Alcohol percentage",
            y= "Count")) +
  coord_cartesian(ylim = c(0, 200))
grid.arrange(plot1, plot2, ncol=2)
```
```{r echo=FALSE}
summary(subset(wine_data$alcohol, wine_data$quality > 5))
```
```{r echo=FALSE}
summary(subset(wine_data$alcohol, wine_data$quality <= 5))
```

Compared the alcohol percentage of highest quality wine and the lowest.
To compare 2 plots I used coord_cortesian function so that my both Y axis would
be the same height. From the plots I could gues that lower quality wines have a
higher chance of having higher alcohol percentage. I got that higher quality
wines have higher alcohol values than lower quality wines. But the lower quality
wine has the max value for wine of 14.9 %.

Could there be any relationship between volatile acidity and quality?
```{r echo=FALSE, message=FALSE, warning=FALSE}
plot3 <- qplot(data = subset(wine_data, quality > 5), x = volatile.acidity) +
  geom_histogram(fill="white",colour="black") +
  labs(list(title = "Wines with quality lower than 5", x = "volatile.acidity",
            y = "Count")) +
  coord_cartesian(ylim = c(0, 130))

plot4 <- qplot(data = subset(wine_data, quality <= 5), x = volatile.acidity) +
  geom_histogram(fill="white",colour="black") + 
  labs(list(title = "Wines with quality greater than 5", 
            x = "volatile.acidity",
            y = "Count")) +
  coord_cartesian(ylim = c(0, 130))
grid.arrange(plot3, plot4, ncol=2)
```
```{r echo=FALSE}
summary(subset(wine_data$volatile.acidity, wine_data$quality > 5))
```
```{r echo=FALSE}
summary(subset(wine_data$volatile.acidity, wine_data$quality <= 5))
```

I can see that the better quality wines ten to be less acidic. I can also see
that lower quality wine has an outlier which increases the x axis. Since the 
3rd quartile is 0.68 and the max value is 1.58 I would asume that theres a
pretty big outlier there. 


# Univariate Analysis
### What is the structure of your dataset?

There are 1599 wines(observations) in the dataset, with 13 variables: 

 * X
 * fixed.acidity
 * citric.acid  
 * residual.sugar
 * chlorides 
 * free.sulfur.dioxide 
 * total.sulfur.dioxide
 * density             
 * pH                  
 * sulphates          
 * alcohol           
 * quality
 
There are no ordered factor variables in this data set.

### What is/are the main feature(s) of interest in your dataset

My main interest in this data set is quality. I want to figure out what variable
influance the quality of wine the most.

### What other features in the dataset do you think will help support your
investigation into your feature(s) of interest?
 
The other features that will help support my investigation will be alcohol
percentage, acidity and pH.

### Did you create any new variables from existing variables in the dataset?

No

### Of the features you investigated, were there any unusual distributions? Did
you perform any operations on the data to tidy, adjust, or change the form of 
the data? If so, why did you do this?

When I was ploting 2 plots side by side of alcohol percentage, the Y axis of 
both of these histograms were different height. To better compare the 2 
histograms I used coord_cartesian(ylim = c(0, 130)) to equalize the heights of
axis.

# Bivariate Plots Section


```{r echo=FALSE, message=FALSE, warning=FALSE}

ggcorr(wine_data,label = TRUE,hjust = .9, size = 3, label_size = 3, 
       label_round = 2, label_alpha = TRUE)
     
```

First I checked for correlations I use cor function and check correlation 
between each of the variables.

I can now see that the quality has the strongest correlation between these 
variables: volatile.acidity, sulphates and alcohol. I was right about that
alcohol and volatile acidity could be the biggest factors for change in quality,
i was wrong about pH though.


```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = alcohol, y = quality), data = wine_data) +
  geom_jitter(alpha = 1/5)  

```
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine_data %>%
  group_by(quality, alcohol) %>%
  summarize())

cor(wine_data$alcohol, wine_data$quality)
```

Checked the correlation between these variables.

I used a scaterplot to see a relationship between alcohol and quality.

In this plot I can see some points that are above 8. I run the max() on quality
variable to check for max value.

I get that max value for quality is 8, but my scatter plot shows some points 
that are above 8. Later I figured out that geom_jitter scatters the points and
when running geom_point it shows the right values.

Transforming the y axis to show median and means.

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = alcohol, y = quality), data = wine_data) +
  geom_jitter(stat = 'summary', fun.y = median)  

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

ggplot(aes(x = alcohol, y = quality), data = wine_data) +
  geom_jitter(stat = 'summary', fun.y = mean)  

```

```{r echo=FALSE}

ggplot(aes(x = alcohol, y = quality), data = wine_data) +
  geom_line(stat = 'summary', fun.y = mean)  

```

The plots show some correlation. Increasing the alcohol percentage might 
increase the quality. But I might be wrong.

Plotting boxplots.

```{r echo=FALSE}
ggplot(aes(x = quality, y = alcohol), data = wine_data) +
  geom_boxplot(aes(group=quality))  

```
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine_data %>%
  group_by(quality, alcohol) %>%
  summarize())

cor(wine_data$alcohol, wine_data$quality)

```

This boxplot also shows that wines with higher alcohol percentage is rated 
better in quality.

Checking quality by volatile.acidity.

```{r echo=FALSE}

ggplot(aes(x = volatile.acidity, y = quality), data = wine_data) +
  geom_jitter(alpha = 1/5)  

```
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine_data %>%
  group_by(quality, volatile.acidity) %>%
  summarize())

cor(wine_data$volatile.acidity, wine_data$quality)

```

The scatterplot seems a bit linear. I want to scatter it a bit more to see the
trends, so I will calculate and apply means to Y axis.

```{r echo=FALSE}

ggplot(aes(x = volatile.acidity, y = quality), data = wine_data) +
  geom_point(stat = 'summary', fun.y = mean)  

```

I can see a trend that as the volatile acidity increases, the quality decreases.

Plotting boxplots.

```{r echo=FALSE}

ggplot(aes(x = quality, y = volatile.acidity), data = wine_data) +
  geom_boxplot(aes(group=quality))

```

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(wine_data %>%
  group_by(quality, volatile.acidity) %>%
  summarize())

cor(wine_data$volatile.acidity, wine_data$quality)

```

From these boxplots I can see the trend that the quality decreases if acidity
increases. I can see some outliers here which might increase the means. I will
use summary to check for the max and mean values.

I can see some outliers, so i created a plot without them.

```{r echo=FALSE, message=FALSE, warning=FALSE}


ggplot(aes(x = quality, y = volatile.acidity), data = wine_data) +
  geom_boxplot(aes(group=quality), outlier.shape = NA) +
  scale_y_continuous(limits = quantile(wine_data$volatile.acidity, c(0.1, 0.9)
                                       ))

```

                                       
```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(quantile(wine_data$volatile.acidity, c(0.1, 0.9)))

cor(wine_data$volatile.acidity, wine_data$quality)

```

I added outlier.shape = NA to remove outlier shapes from the plot and then set 
the both bottom and upper limits of y axis. This plot also shows this trend that
the higher the acidity - the lower is quality. I summarised only the values
without outliers using quantile function. I get min value of 0.31 and max value
of 0.75

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the 
investigation. How did the feature(s) of interest vary with other features in 
the dataset?

I checked my feature of interest (quality) against 2 variables: 
volatile.acidity and alcohol. The relationship with alcohol was that I saw a
trend that higher alcohol percentage means higher quality. For volatile acidity
I saw that high acidity means lower quality.

### Did you observe any interesting relationships between the other features 
(not the main feature(s) of interest)?

I only observed the relationships to quality, so no i did not observe 
interesting relationships between other features.

### What was the strongest relationship you found?

In my main feature of interest I found out that quality has a strong 
relationship with alcohol percentage. As the alchohol percentage increases,
so does quality of the wine. The correlation coeficent between these 2 variables
is 0.48. The other strong relationship I found was between acids and pH which
is not surprising at all.

# Multivariate Plots Section

##14. Creating a new summary

```{r echo=FALSE, message=FALSE, warning=FALSE}


new_data <- wine_data %>%
  group_by(volatile.acidity, quality, pH, alcohol) %>%
  summarize()

summary(new_data)

```

I created a group of variables to easier summarise the data that I want.

##14. Checking various plots.
```{r echo=FALSE}

ggplot(aes(x = volatile.acidity, y = pH), data = new_data) +
  geom_jitter() +
  facet_wrap( ~ quality)

```
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine_data %>%
  group_by(volatile.acidity, quality, pH) %>%
  summarize())
cor(wine_data$volatile.acidity, wine_data$pH)
```

I wanted to see if there are some anomalies in residual.sugar vs pH 
relationship. I created a plot for each of the quality score in hoping to see
something unusual. I can see that all of these plots has similar clustering and
not much stands out. I can see a small trend that higher quality wines tend to
be stronger.

```{r echo=FALSE}
ggplot(aes(x = volatile.acidity, y = alcohol, color = factor(quality)),
       data = wine_data) +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine_data %>%
  group_by(volatile.acidity, quality, alcohol) %>%
  summarize())
cor(wine_data$volatile.acidity, wine_data$alcohol)
```

From this plot I can see that the highest quality wines tend to heave more
alcohol in it and it also is from low to medium acidity. Lower quality wines
usualy have a wider range of acidity maxing at 1.6, it also tend to have less 
alcohol in it. The weird thing is that as the best quality has the most alcohol,
lowering the quality drasticaly reduces the alcohol percentage, and then at the
lowest point in quality, the alcohol percentage drasticaly increases.


```{r echo=FALSE}
  
ggplot(aes(x = volatile.acidity, y = pH, color = factor(quality)),
       data = wine_data) +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))


```
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine_data %>%
  group_by(volatile.acidity, quality, pH) %>%
  summarize())
cor(wine_data$volatile.acidity, wine_data$pH)
```

From this plot I can see that highest quality wines tend to have slightly higher
level of ph while retaining medium range of volatile acidity. Lower quality 
wines tend to have more acidity and a little bit lowe pH level than highest
quality wine.

```{r echo=FALSE}
ggplot(aes(x = volatile.acidity, y = quality), data = wine_data) +
  geom_jitter(aes(color = pH)) +
  geom_smooth(method = 'lm', color = 'Red')
```

I can see that the higher pH levels might be lowering the quality, maybe 
because high pH means more acidity.


```{r echo=FALSE, message=FALSE, warning=FALSE}
  
ggplot(aes(x = volatile.acidity, y = quality), data = wine_data) +
  geom_line(aes(color = pH), stat = 'summary', fun.y = mean) + 
  geom_smooth()

```
```{r echo=FALSE}
summary(wine_data %>%
  group_by(volatile.acidity, quality, pH) %>%
  summarize())
cor(wine_data$volatile.acidity, wine_data$quality)
```

This further tells me that there is correlation between quality and volatile 
acidity.
```{r echo=FALSE}

ggplot(aes(x = quality, y = alcohol), data = wine_data) +
  geom_jitter(aes(color = volatile.acidity)) +
  geom_smooth(method = 'lm', color = 'Red')

```
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine_data %>%
  group_by(volatile.acidity, quality, alcohol) %>%
  summarize())
cor(wine_data$volatile.acidity, wine_data$quality)
```

I can clearly see that increasing alcohol percentage we would increase quality,
but I see no big trends petween volatile.acidity and these 2 variables. Maybe
I can see something if I compare alcohol, quality and pH?

```{r echo=FALSE}

ggplot(aes(x = quality, y = alcohol), data = wine_data) +
  geom_jitter(aes(color = pH)) +
  geom_smooth(method = 'lm', color = 'Red')
```
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(wine_data %>%
  group_by(alcohol, quality, pH) %>%
  summarize())
cor(wine_data$alcohol, wine_data$quality)
```

I can see a small correlation between pH and alcohol, since lower pH wines tend 
to sit in the bottom of the plot, and that means lower alcohol could mean lower
pH value.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the 
investigation. Were there features that strengthened each other in terms of 
looking at your feature(s) of interest? 

I was not surprised with the outcomes
of my plots. I suspected that volatile acidity lowers the quality and higher 
alcohol concentration might increase the quality. Checking the correlation
between these variables showed me that they are correlated, and the plots
helped me to be sure about that.

### Were there any interesting or surprising interactions between features?

Before my analysis I was really thinking that lower alcohol percentage would
mean better quality, but that proved me wrong. I was also thinking that
pH and volatile acidity should have very similar results for highest
quality wines, but it seems that its not. Another interesting thing was that
the lower quality wines has less and less alcohol in it, bus as the quality 
reaches lowest point, the alcohol percentage drasticaly increases.

### OPTIONAL: Did you create any models with your dataset? Discuss the  \
strengths and limitations of your model.

No.

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One, message=FALSE, warning=FALSE}

ggplot(aes(x = quality, y = volatile.acidity), data = wine_data) +
  geom_boxplot(aes(group=quality), outlier.shape = NA, fill = I('#F79420')) +
  scale_y_continuous(limits = quantile(wine_data$volatile.acidity, c(0.1, 0.9)
                                       )) +
  labs(title = "Volatile acidity relationship with quality",
       x = 'Quality of wine', y = 'Amount of acetic acid in wine')


```

### Description One
I wanted to see what is the relationship between acidity and quality. I was
hoping for this relationship to be somehow neutral. To see the relationship 
I plotted a boxplot. The boxplot had many outliers and was interfering with 
my means, so I had to remove the outliers to see the real relationship. After
plotting the final boxplot I see clear tendency that increasing 
acidity - decreases the quality.

### Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(aes(x = volatile.acidity, y = alcohol, color = factor(quality)),
       data = wine_data) +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality')) +
  labs(title = "Volatile acidity relationship with alcohol and quality",
       x = 'Volatile acidity g / dm^3', y = 'Alcohol %')


```

### Description Two

Interpreting this plot I found out that the highest quality wine has the most
alcohol in it. As the quality decreases, so does alcohol percentage and volatile
acidity. Reaching lowest point, the alcohol percentage drasticaly jumps higher
at higher levels of volatile acidity which I find strange.

### Plot Three
```{r echo=FALSE, Plot_Three}

ggplot(aes(x = volatile.acidity, y = pH, color = factor(quality)),
       data = wine_data) +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE,size=1) +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality')) +
  labs(title = "Volatile acidity relationship with pH value and quality",
       x = 'Volatile acidity g / dm^3', y = 'pH value')

```

### Description Three
From this plot I can see that highest quality wines tend to have slightly higher
level of ph while retaining medium rango of volatile acidity. Lower quality 
wines tend to have more acidity and a little bit lowe pH level than highest
quality wine.

In this plot I noticed that the highest quality wines has higher level of pH and
lower levels of volatile acidity, which is weird because I thought that
increasing the pH should also increase the volatile acidity. As the quality 
goes down, the volatile acidity levels increase and pH levels stays lower that
in highest quality wines.



# Reflection
The Red Wine Quality dataset includes 1599 observations with 13 variables. As 
I was exploring the data set I found that the most interesting variable for me
is quality, which I based my project on. I Tried to check pH, alcohol, acidity
variables against quality to see for some interesting correlations.  

I strugled with some of the outliers screwing my boxplots, but later I found
out how to set the limits on Y axis using quantile functiont by removing top 
10% and bottom 10% of data from the plot. Another difficulty was that the 
dataframe is tiny compared to others, and 1599 observations doesn't make
beutifull scatterplots. Besides these problems all went well and I noticed 
that the biggest correlations between quality was alcohol and acidity. 
Higher alcohol improves the quality while acidity decreases it.

For better analysis in the future I would try to test all of the variables 
and maybe join the white wine and red wine datasets to see evem bigger trends.

